PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-04-19 papers/NAR_web_server_2009/paper.sup.tex
Adding tex version of sup. materials

@@

\newlabel{sensitivityFig}{1}

@@

\title{PEACE: {\underline P}arallel {\underline E}ST {\underline A}nalysis
  and {\underline C}lustering {\underline E}ngine \\Supplementary
  Materials}

@@

Once the MST is has been calculated, our last step is to remove all
edges exceeding a threshold weight $T$, taking the resulting
components as our clusters.  \textsc{wcd}, when faced with a similar
challenge, sets the threshold at $T=40$, hypothesizing that EST pairs
with a $d^2$ distance of greater than 40 are unlikely to overlap.  We
have set out threshold to $T=130$.

@@

In {\it Hazelhurst et al.} [\cite{Hazelhurst08a}] the authors conduct
an investigation of \textsc{wcd} against a number of clustering
tools.  As they make a convincing argument that \textsc{wcd} is returning
better results then the tools against which they compare, we limit our
analysis to a comparison of \textsc{peace} against \textsc{wcd} and the \textsc{cap3} assembly
tool [\cite{Huang99}] -- a tool which implicitly clusters while
performing assembly.  We do so by applying all three tools to a number
of data sets, both simulated and real.

@@

While our primary measurement of tool result is sensitivity
(Figure~\ref{sensitivityFig}), a number of alternative measurements
were investigated as well. The {\it Jaccard} index measures both
sensitivity and specificity, calculated as the ratio of true positives
to the sum of true positives, true negatives, and false positives ($tp
/ (tp+fn+fp)$) [\cite{Hazelhurst08a}].  Notably, when we removed duplicated genes before
generating the simulated EST sets, we found all three tools to have
100\% specificity: no false positives.  In such a case the Jaccard
Index is identical to the Sensitivity Index, hence \textsc{peace} exhibits the
same improvement on both indexes relative to \textsc{wcd} and \textsc{cap3}.  Applying
the tools to the original genes (in which 21\% of the genes were
highly similar to genes in the remaining 79\%), we find only a slight
difference in the Jaccard Index, and the same relative performance of
the three tools (Figure~\ref{JiSe}).  The sensitivity of each tool to
duplications is discussed below.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.5]{4JiSe.eps}
\label{JiSe}
}
\caption{Comparisons of Sensitivity and the Jaccard Index as a
  function of base read error rate, based on 30 simulated EST sets
  derived from a set of 79 genes with no significant similarity (left
  column), and an additional 21 genes showing significant similarity
  to the base set (right column).  Blue/Solid = \textsc{peace}, Green/Dash =
  \textsc{wcd}, Black/Dot-Dash = \textsc{cap3}; vertical tics = 95\% confidence
  intervals on estimates.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

Type 1 and Type 2 errors, as defined in {\it Wang et al.}
[\cite{Wang04}], measure quality at the level of the genes from which
the ESTs were derived.  For Type 1 error, we look at the fraction of genes
that were  broken into two or more partitions, while in
Type 2 error we look at the fraction of clusters that contain two or
more genes.   In Figure~\ref{t1t2} we compare the Type 1 and Type 2 error rates of
the three tools for varying levels of simulated base read error.  For
Type 1 error we find that while all tools do equally poorly when the
EST is subjected to more than a $5\%$ error rate, \textsc{peace} does
significantly better than \textsc{wcd} in the lower ranges (while
\textsc{wcd} does correspondingly better than \textsc{cap3}).  When
looking at Type 2 error rate, we find that \textsc{peace} does a
generally better job than \textsc{cap3} in separating genes into
different clusters.  We note, however, that Type 1 error is the more
important measurement: genes incorrectly split between clusters cannot
be later recovered (without re-examining the cluster as a whole), but
there is the potential for an assembly program to later identify
incorrectly merged clusters based only on local cluster information.

%%%%%%%%%%%%%%
\begin{figure}[tbp]
\centerline{\includegraphics[scale=0.5]{t1t2.eps}
\label{t1t2}
}
\caption{Type 1 error rate (fraction of genes partitioned) and Type 2
  error rate (fraction of clusters containing multiple genes) as a
  function of base read error rate.  Estimates averaged over
  application to 30 simulated ESTs sets derived from the full 100
  zebra fish gene set. Blue/Solid = \textsc{peace}, Green/Dash =
  \textsc{wcd}, Black/Dot-Dash = \textsc{cap3}; vertical tics = 95\%
  confidence intervals on estimates (not shown for Type 1 error due to
  the extremely small variance in estimates).}
\end{figure}
%%%%%%%%%%%%%%

@@

\begin{figure}[tbp]
\centerline{
\label{dups}
\includegraphics[scale=0.35]{duplicates.eps}
}
\caption{Ability to distinguish duplicates as a function of
  divergence.  Estimates averaged over 30 trials; one trial consists
  of taking a random gene, copying it and stochastically changing bases
  at the specified rate, then using the two genes as the bases for
  generating a simulated set. Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3};
  variance was too small for visible plotting of
  confidence intervals.}
\end{figure}

@@

%%%%%%%%%%%%%
\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.35]{seq_time.eps}
\label{seq_runtime}
}
\caption{Single Processor Runtime as a function of input size,
  averaged over 50 simulated EST sets; Blue/Solid = \textsc{peace},
  Green/Dashed = \textsc{wcd}; vertical tics represent= 95\% confidence
  intervals on estimates.}
\end{figure}
\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.5]{par_time.eps}
\label{par_runtime}
}
\caption{Runtime of a function of input size when holding constant the
  ratio of set size to number of processors used; averaged over 30
  simulated EST sets.  Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd};
  vertical tics represent= 95\% confidence intervals on estimates.}
\end{figure}
%%%%%%%%%%%%%

\section{Applications to Real Data Sets}

In Table~\ref{chlamy} we lay out more comprehensive results for the
application of our tool to an EST set collected from the {\it
  Chlamydomonas reinhartdii} genome, while in Table~\ref{mouse} 
we see the results for the mouse set used for the \textsc{wcd} analysis
[\cite{Hazelhurst08a}].  We note that for the Chlamydomonas genome the
correct clustering is not known; our estimates of sensitivity are
based on as assignment of ESTs to the sequenced genomes by the Gmap tool
[\cite{Wu05}], leaving some potential for error in the  reference.
Further, we found a number of ligations in the data (ESTs derived by
incorrectly combining multiple transcripts),  which were removed before testing.

Interestingly, where in the simulated results we saw significant
improvement in \textsc{peace} over \textsc{wcd} in terms of result quality and only a
light improvement in runtime, here we see the reverse. The two tools are
roughly comparable  in terms of sensitivity, Type 1 error and Type 2
error, while \textsc{peace} shows a 10\% and 20\% improvement in runtime (both
showing significant improvements over \textsc{cap3}).  Predictably , the full
assembly tool does somewhat better in terms of specificity (the Jaccard
Index and Type 2 error), though scores here are worse for all three
tools due to the increased number of homologous gene pairs as compared
to our simulated sets.

\begin{table}[tbp]
\begin{center}
\begin{tabular} {| c || c | c | c | c | c | c | c | c |}
\hline
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} & \begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & Singletons & \begin{tabular}{c} Single
      \\ processor \\ runtime (s) \end{tabular}\\
\hline \hline \textsc{peace} & 0.958 & 0.386 & 0.623 & 0.026 & 19649 & 10493 & 8799 \\
\hline \textsc{wcd} & 0.94 & 0.500 & 0.654 & 0.021 & 22433 & 13000 & 9913 \\
\hline \textsc{cap3} & 0.77 & 0.740 & 0.737 & 0.015 & 33771 & 22169 &  \\
\hline
\end{tabular}
\label{chlamy}
\end{center}
\caption{{\bf Chlamydomonas reinhartdii}: 189975 ESTs, average length
  553 bp, estimated 9886 actual genes represented.}
\end{table}

\begin{table}[tbp]
\begin{center}
\begin{tabular} {| c || c | c | c | c | c | c | c | c |}
\hline
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} & \begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & Singletons & \begin{tabular}{c} Single
      \\ processor \\ runtime (s) \end{tabular}\\
\hline \hline \textsc{peace} & 0.936 & 0.348 & 0.340 & 0.034 & 18338 & 8206 & 820\\
\hline \textsc{wcd} & 0.932 & 0.476 & 0.350 & 0.0275 & 18787 & 8554 & 1000\\
\hline \textsc{cap3} & 0.830 & 0.805 & 0.486 & 0.0135 & 25040 & 14925 & \\
\hline
\end{tabular}
\caption{{\bf A076951 (Mouse EST set)}: 76914 ESTs, average length 427,
  13240 actual genes represented.}
\label{mouse}
\end{center}
\end{table}

>>>>

\newcommand{\peace} {{\small PEACE}}
\newcommand{\wcd} {{\small WCD}}
\newcommand{\capthree} {{\small Cap3}}
\newcommand{\easycluster} {{\small EasyCluster}}

@@

\newlabel{sensitivityFig}{3}
\newlabel{SangerTable}{1}

@@

\title{PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly
  and {\underline C}lustering of Gene {\underline E}xpression}

@@

Once the MST is has been calculated, our last step is to remove all
edges exceeding a threshold weight $T$, taking the resulting
components as our clusters.  \textsc{wcd}, when faced with a similar
challenge, sets the threshold at $T=40$, hypothesizing that EST pairs
with a $d^2$ distance of greater than 40 are unlikely to overlap.  \mc{We
have set our threshold to the same $T=40$ (with discussion below on
the effects of changing this).}

\subsection{Adaptive $d^2$}

In the above discussion of $d^2$ we assume sequences of at least 100 bp, and
in pratice use a window of size 100.  While the technique should be,
in principle, adaptable to short-read sequences (decreasing window
size as necessary), there are some problems in the details.
Diminishing window sizes implies both diminishing accuracy (requiring
the thrshold be adjusted) and increasing runtime (as there are
potenitally more window pairs).  More importantly, there is the
problem of non-homogonous sequence lengths: short-read technologies
tend to produce short reads, but can still include quite long reads in
their output.  Hence it is not immediately clear how to compare a very
short read against a very long one.

We address these issues with our {\it adaptive $d^2$} strategy that
works as follows: we put sequences, based on length, into three {\it
  overlapping} groups (i.e. sequences less than length XXX, sequences
between length YYY and ZZZ, and seqeunces greater than AAA).  For each
group we pick a constant windowsize and threshold tailored to the
length range of the group (as well as filtering heurstic constants),
and use those values when computing the $d^2$ distance for two
sequences in the same group.  We do not compute distances between
sequences in different groups. Once computed, we assign to the MST
edge the ratio of the $d^2$ score to the group threshold, thus
normalizing the weight for comparisons ``between groups'' and
effectivally giving us an edge-removal threhold of 1.

By restricting $d^2$ comparisons to within a group, we minimize the
effets of the hetrogenous sequences lengths by restricting our
comparisons to only sequences of similar lengths.  By not restrcicting
ourselves to a single window size, we allow ourselves to, when possible, avoid the
preformance hit of using a small window.  As we are not comparing
sequences from different groups we bypass the question of how to
compare a largelong sequence against a short one.  And because the
groups overlap in content, we still get implicit comparisons between
long and short sequences in terms of the resulting path of the minimum
spanning tree that makes use of these ``linking'' nodes.

In the \textsc{peace} implementation, we subdevide into three groups:
NEED INFO.

@@

In {\it Hazelhurst et al.} [\cite{Hazelhurst08a}] the authors conduct
an investigation of \textsc{wcd} \mc{(version 0.5.1)} against a number of clustering
tools.  As they make a convincing argument that \textsc{wcd} is returning
better results then the tools against which they compare, we limit our
analysis to a comparison of \textsc{peace} against \textsc{wcd} and the \textsc{cap3} assembly
tool [\cite{Huang99}] -- a tool which implicitly clusters while
performing assembly.  We do so by applying all three tools to a number
of data sets, both simulated and real.

@@

\mc{ Our primary measurement tools if mesurement are sensitivity, the
Jaccard Index, Type 1 error and Type 2 error
(Figure~\ref{sensitivityFig}).  Sensitivity is the ratio for true
positives to actual positives ($tp / (tp + fn)$) -- the fraction of
sequences from the same transcript that are identified as such.  The
{\it Jaccard} index combines both sensitivity and specificity,
calculated as the ratio of true positives to the sum of true
positives, true negatives, and false positives ($tp / (tp+fn+fp)$)
[\cite{Hazelhurst08a}].  Notably, when we removed duplicated genes
before generating the simulated EST sets, we found all three tools to
have 100\% specificity: no false positives. }

\mc{Type 1 and Type 2 errors, as defined in {\it Wang et al.}
[\cite{Wang04}], measure quality at the level of the genes from which
the ESTs were derived.  For Type 1 error, we look at the fraction of genes
that were broken into two or more partitions, while in
Type 2 error we look at the fraction of clusters that contain two or
more genes.   }

\mc{A fifth metric is to look at the number of singletons produce by a
  tool:  sequences that  could not  be  assigned to  any cluster.   In
  practice such a sequence could represent compelte coverage of a very
  short transcript (one that may be otherwise had to identify), but is
  more likely to be an error in the clustering algorithm.  In order to
  identify  the first  it is  important  to minimize  the second.   In
  Figure~\ref{singletons} we look at the number of singletons produced
  by each tool when run on  simulated data in which there should be no
  singletons produced,  seeing a very slight improvment  of PEACE over
  WCD in certain error ranges.}

\begin{figure}[tbp]
\centerline{
\label{dups}
\includegraphics[scale=0.35]{pics.d/singletons_40.pdf}
}
\caption{Number of singletons produced by each tool in a simulated run
  where a fully correct clustering would have no singletons.  Values
  average over 30 triles.  Blue/Solid = \textsc{peace}, Green/Dashed =
  \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3}.}\label{singletons}
\end{figure}

@@

\begin{figure}[tbp]
\centerline{
\label{dups}
\includegraphics[scale=0.35]{pics.d/duplicates_40.pdf}
}
\caption{Ability to distinguish duplicates as a function of
  divergence.  Estimates averaged over 30 trials; one trial consists
  of taking a random gene, copying it and stochastically changing bases
  at the specified rate, then using the two genes as the bases for
  generating a simulated set. Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3};
  variance was too small for visible plotting of
  confidence intervals.}
\end{figure}

\subsection{Raising the Threshold}
\mc{ 
  The \wcd\/ tool uses a $d^2$ threshold of $40$ to determine whether
  two sequences are in the same cluster.  As this is equivilent to our
  $T$ threshold for dismissing MST edges, we have set it to the same
  default value when used for Sanger sequences.  However, we did
  experiment with higher values, to interesting results.  A basic
  analysis indicated that a $T$ value as high as 130 should not have a
  significant impact on its effectivness as a descriminator.  And in
  fact this was born out in simulation, as shown in
  Figure~\label{SiJiT1T2130}.  In contrast with the equivilent results
  for a threshold of $T=40$, as shown in the manuscript, we see
  considerable improvement in \peace\/ over \wcd\/ in all metrics
  except Type 2 error, in which \wcd\/ slightly
  better. Unsurprisingly, this also makes a significant difference in
  the tool's ability to distinguish duplications: we see fairly poor
  performance in Figure~\ref{dups130}.  
}

\mc{
  Most importantly, while the higher threshold appeared to be better
  in theory and in analysis, it did not hold in practice.  In
  Table~\ref{SangerTable130} we show the match to the manuscript
  Table~\ref{SangerTable}, and observe the poor performance of
  \peace\/ when looking at the Jaccard index.  It is clear that in the
  real data, the level of similarity is consierably higher then in the
  simulations, thus making the reduced specificity of the higher
  threshold considerably more of a problem.
XXX}

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
  \caption{Comparisons of Sensitivity, Jiccard Index, Type 1 error and
    Type 1 error, based on the
    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
    fish genes  (see
    Supplementary Materials, Section C, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
    tics = 95\% confidence intervals on estimates.  Intervals are not
    presented for Type 1 error due to the effective lack of
    variance.}\label{SeJiT1T2130}
\end{figure}

\begin{figure}[tbp]
\centerline{
\label{dups130}
\includegraphics[scale=0.35]{pics.d/duplicates_130.pdf}
}
\caption{Ability to distinguish duplicates as a function of
  divergence.  Estimates averaged over 30 trials; one trial consists
  of taking a random gene, copying it and stochastically changing bases
  at the specified rate, then using the two genes as the bases for
  generating a simulated set. Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3};
  variance was too small for visible plotting of
  confidence intervals.}
\end{figure}

@@

\mc{All tests were run on the University of Miami Redhawk cluster,
  made up of 3.2 GHz Intel Xeon EM64T CPUs, each with a 2MB cache and
  800 MHz side bus, model number Xeon LV 3.0 (2005).}

%%%%%%%%%%%%%
\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.35]{pics.d/seq_time.pdf}
\label{seq_runtime}
}
\caption{Single Processor Runtime as a function of input size,
  averaged over 50 simulated EST sets; Blue/Solid = \textsc{peace},
  Green/Dashed = \textsc{wcd}; vertical tics represent= 95\% confidence
  intervals on estimates.}
\end{figure}
\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.5]{pics.d/par_time.pdf}
\label{par_runtime}
}
\caption{Runtime of a function of input size when holding constant the
  ratio of set size to number of processors used; averaged over 30
  simulated EST sets.  Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd};
  vertical tics represent= 95\% confidence intervals on estimates.}
\end{figure}
%%%%%%%%%%%%%

\section{Raising the Threshold Real Data Sets}

\begin{table}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular} \\
\hline EasyCluster Human & \peace & 0.999 & 0.106 & 0.027 & 0.030 & 67 & 2 & 295 \\
\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\hline
\hline WCD A076941 & \peace & 0.937 & 0.345 & 0.337 & 0.0342 & 18256 & 8135 & 1293 \\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\label{SangerReal130}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
  standard quality measurments with $T=130$.}
\end{table}
